---
title: "Lecture 4: Wins above Replacement I"
format: html
---


## Overview

### Key Topics

Expected runs framework





## Obtaining pitch-by-pitch data


### PitchF/X and StatCast
<!--
  What is PitchF/X and What is StatCast
-->

### Accessing data from StatCast

Major League Baseball hosts a public-facing web interface for accessing StatCast data.
Using that interface, users can pull up data for individual players or about all pitches of a certain type.
Powering this website is an application programming interface (API), which allows software applications to connect to the underlying StatCast database.
It is through this API that the **baseballR** package acquires data.
If you have not yet installed that package, you can do so using the following code
```{r install-statcast, eval=FALSE}
devtools::install_github(repo = "BillPetti/baseballr")
```
The **baseballR** package provides a function [`baseballr::statcast_search()`](https://billpetti.github.io/baseballr/reference/statcast_search.html) that allows users query all StatCast data by date, player, or player type.
One of the original StatCast authors, Bill Petti, wrote a wrapper function that uses `baseballr::statcast_search()` to pull down an entire season's worth of pitch-by-pitch data; see [this blog post](https://billpetti.github.io/2021-04-02-build-statcast-database-rstats-version-3.0/) for more the wrapper function code and [this earlier post]() for details about its design.
Since he published his original function, StatCast has added some new fields, necessitating a few changes.
The code below defines a new scraper, which we will use in the course. An R script containing this code is available at []().
At a high level, the scraping function pulls data from StatCast on a week-by-week basis. 

```{r annual-statcast-query-def, eval = FALSE}
#| code-fold: true
#| code-summary: "Show the code"
annual_statcast_query <- function(season) {
  
  data_base_column_types <- 
    read_csv("https://app.box.com/shared/static/q326nuker938n2nduy81au67s2pf9a3j.csv")
  
  dates <- 
    seq.Date(as.Date(paste0(season, '-03-01')),
             as.Date(paste0(season, '-12-01')), 
             by = '4 days')
  
  date_grid <- 
    tibble::tibble(start_date = dates, 
                   end_date = dates + 3)
  
  safe_savant <- 
    purrr::safely(scrape_statcast_savant)
  
  payload <- 
    purrr::map(.x = seq_along(date_grid$start_date),
               ~{message(paste0('\nScraping week of ', date_grid$start_date[.x], '...\n'))
                 payload <- 
                   safe_savant(start_date = date_grid$start_date[.x], 
                               end_date = date_grid$end_date[.x], 
                               type = 'pitcher')
                 return(payload)
                 })
  
  payload_df <- purrr::map(payload, 'result')
  
  number_rows <- 
    purrr::map_df(.x = seq_along(payload_df),
                  ~{number_rows <- 
                    tibble::tibble(week = .x, 
                                   number_rows = length(payload_df[[.x]]$game_date))
                  }) %>%
    dplyr::filter(number_rows > 0) %>%
    dplyr::pull(week)
  
  payload_df_reduced <- payload_df[number_rows]
  
  payload_df_reduced_formatted <- 
    purrr::map(.x = seq_along(payload_df_reduced), 
               ~{cols_to_transform <- 
                 c("pitcher", "fielder_2", "fielder_3",
                   "fielder_4", "fielder_5", "fielder_6", "fielder_7",
                   "fielder_8", "fielder_9")
               df <- 
                 purrr::pluck(payload_df_reduced, .x) %>%
                 dplyr::mutate_at(.vars = cols_to_transform, as.numeric) %>%
                 dplyr::mutate_at(.vars = cols_to_transform, function(x) {ifelse(is.na(x), 999999999, x)})
               character_columns <- 
                 data_base_column_types %>%
                 dplyr::filter(class == "character") %>%
                 dplyr::pull(variable)
               numeric_columns <- 
                 data_base_column_types %>%
                 dplyr::filter(class == "numeric") %>%
                 dplyr::pull(variable)
               integer_columns <- 
                 data_base_column_types %>%
                 dplyr::filter(class == "integer") %>%
                 dplyr::pull(variable)
               df <- 
                 df %>%
                 dplyr::mutate_if(names(df) %in% character_columns, as.character) %>%
                 dplyr::mutate_if(names(df) %in% numeric_columns, as.numeric) %>%
                 dplyr::mutate_if(names(df) %in% integer_columns, as.integer)
               return(df)
               })
  
  combined <- payload_df_reduced_formatted %>%
    dplyr::bind_rows()
  
  return(combined)
}
```


<!--
  1. How to scrape pitch-by-pitch data from StatCast
  2. How to get plate appearances? Filter down to the smallest 
  3. Clean up certain fields
-->


## Expected Runs

<!--
  Definition, based on the 24 combinations of outs and baserunners
  Build the matrix
  Add EPA start and EPA after for each plate appearance. Also get number of runs scored in that plate appearance
-->


### Adjusting for platoon advantages & park effects



<!--
  Regress delta onto park and indicator Bhand == Phand
  Get those residuals and add them to the data frame
-->


## Base-running

Quote from Baumer et al. "baserunners should only get credit for advancement beyond what would be expected given their starting locations, number of outs, and the hitting event that occurred"
<!--
  Regress residuals (epsilon) from above onto indicators for gamestate; hitting event that occured
  Resulting residual (eta): represent the portion of the adjusted oﬀensive run value that is attributable to the baserunners.
-->
### Expected base advancement

## Batting

<!--
eta: portion of adjusted offensive run value attributable to baserunners.
mu = epsilon-eta attributable to the hitter
** need to compare batter relative to other players at his position **

-->


